---
title: "Beta-GAMM Analysis of Fetal Y-Chromosome Fraction"
output: html_document
date: "2025-09-04"
---

## Objective

This analysis evaluates the nonlinear association of gestational age and maternal BMI with fetal Y-chromosome fraction. A beta generalized additive mixed model accounts for the bounded response and repeated measurements from the same participant.

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(readxl)
library(dplyr)
library(forcats)
library(lubridate)
library(mgcv)
library(ggplot2)

repo_root <- if (basename(getwd()) == "analysis") normalizePath("..") else getwd()
data_path <- file.path(repo_root, "data", "nipt_data.xlsx")
if (!file.exists(data_path)) {
  stop("The authorized dataset is not included. Place it at data/nipt_data.xlsx.")
}
```

## Load and prepare the data

The workbook is expected to contain a worksheet named `male_data` with the English column names listed in `data/README.md`.

```{r data-preparation}
df <- read_excel(data_path, sheet = "male_data") %>%
  transmute(
    maternal_id = factor(maternal_id),
    maternal_age = as.numeric(maternal_age),
    maternal_bmi = as.numeric(maternal_bmi),
    gestational_age = as.numeric(gestational_age),
    y_fraction = as.numeric(y_chromosome_fraction),
    conception_method = factor(conception_method),
    parity = as.integer(parity),
    aneuploidy = as.numeric(aneuploidy)
  ) %>%
  filter(aneuploidy == 0 | is.na(aneuploidy)) %>%
  filter(
    is.finite(maternal_bmi),
    is.finite(gestational_age),
    is.finite(y_fraction),
    !is.na(maternal_id)
  )

if (max(df$y_fraction, na.rm = TRUE) > 1) {
  df <- df %>% mutate(y_fraction = y_fraction / 100)
}

n_total <- nrow(df)
df <- df %>%
  mutate(y_fraction_adjusted = (y_fraction * (n_total - 1) + 0.5) / n_total)
```

## Fit candidate models

```{r model-fitting}
m1 <- gam(
  y_fraction_adjusted ~ s(gestational_age, k = 6) + s(maternal_id, bs = "re"),
  data = df,
  family = betar(link = "logit"),
  method = "REML"
)

m2 <- gam(
  y_fraction_adjusted ~ s(gestational_age, k = 6) + s(maternal_bmi, k = 5) +
    s(maternal_id, bs = "re"),
  data = df,
  family = betar(link = "logit"),
  method = "REML"
)

m3 <- gam(
  y_fraction_adjusted ~ s(gestational_age, k = 6) + s(maternal_bmi, k = 5) +
    te(gestational_age, maternal_bmi, k = c(6, 5)) + s(maternal_id, bs = "re"),
  data = df,
  family = betar(link = "logit"),
  method = "REML"
)

m4 <- gam(
  y_fraction_adjusted ~ s(gestational_age, k = 6) + s(maternal_bmi, k = 5) +
    te(gestational_age, maternal_bmi, k = c(6, 5)) + maternal_age +
    conception_method + parity + s(maternal_id, bs = "re"),
  data = df,
  family = betar(link = "logit"),
  method = "REML"
)
```

## Compare model fit

```{r model-comparison}
model_comparison <- data.frame(
  model = c("m1", "m2", "m3", "m4"),
  AIC = c(AIC(m1), AIC(m2), AIC(m3), AIC(m4)),
  REML_score = c(
    as.numeric(-2 * logLik(m1)),
    as.numeric(-2 * logLik(m2)),
    as.numeric(-2 * logLik(m3)),
    as.numeric(-2 * logLik(m4))
  )
)

model_comparison
anova(m1, m2, m3, m4, test = "Chisq")
```

## Diagnose the selected model

```{r diagnostics, fig.width=7, fig.height=5}
selected_model <- m3
diagnostic_data <- data.frame(
  fitted = fitted(selected_model, type = "response"),
  observed = model.response(model.frame(selected_model)),
  residual = residuals(selected_model, type = "deviance"),
  linear_predictor = predict(selected_model, type = "link")
)

ggplot(diagnostic_data, aes(fitted, observed)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
  geom_point(alpha = 0.55, color = "#2B6CB0") +
  geom_smooth(method = "loess", se = FALSE, color = "#E76F51") +
  labs(
    title = "Observed versus fitted fetal Y-chromosome fraction",
    x = "Fitted fraction",
    y = "Observed fraction"
  ) +
  theme_classic()

ggplot(diagnostic_data, aes(linear_predictor, residual)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_point(alpha = 0.55, color = "#3366CC") +
  geom_smooth(method = "loess", se = FALSE, color = "#E76F51") +
  labs(
    title = "Deviance residuals versus linear predictor",
    x = "Linear predictor",
    y = "Deviance residual"
  ) +
  theme_classic()
```

## Interpretation note

The fitted relationships describe associations in the competition dataset. They should not be interpreted as causal effects or used for clinical decisions without external validation.


